---
title: "Group Project: Covid Origins"
author: "Name: Sal Figueroa  \n Partner: Kylie Stearns"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
#  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---
```{r setup, include = FALSE}

#### Load necessary packages ####
# * These packages are not necessary to complete the assignment and or only used 
#   to provide an example. 
packages <- c("knitr", "kableExtra", "magrittr", "readr", "geosphere")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
#library(kableExtra)
library(magrittr)
library(readr)
library(geosphere)
```

```{r Data setup, include = FALSE}



#time_series_covid19_deaths_global.csv
deaths_global <- as.data.frame(read.csv("https://raw.githubusercontent.com/Figgs0bit/CSIT165-CovidGroupProj/refs/heads/main/time_series_covid19_deaths_global.csv"))

#time_series_covid19_recovered_global.csv
recovered_global <- as.data.frame(read.csv("https://raw.githubusercontent.com/Figgs0bit/CSIT165-CovidGroupProj/refs/heads/main/time_series_covid19_recovered_global.csv"))
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*. 
Your and your partner's role is to play a data scientist from one of these two entities.

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

## Project Objectives

### Objective 1
*Predict where the origin started based on the area with the greatest number of confirmations and deaths on the first recorded day in the data set. Show this is the origin using an if statement.*

```{r ob1}
#find max value in column (10.22.20)
#check to see how many max location there is
#run conditional loop 
max_value <- 0
RowMax <- 0

for(i in 1:nrow(deaths_global))
{
num <- deaths_global[i,5]
  
  if (num > max_value)
    {
    max_value <- num
    RowMax <- i
    }
    
#province.state
Acol <- deaths_global[RowMax,1]
#county.region
Bcol <- deaths_global[RowMax,2]
#latitude
Ccol <- deaths_global[RowMax,3]
#longitude
Dcol <- deaths_global[RowMax,4]
#deaths on x1.22.20 (date)
DeathVal <- deaths_global[RowMax,5]
} 

cat(Acol,", ", Bcol,", ", Ccol,", ", Dcol,", ", DeathVal, "\n")
```

### Objective 2
```{r ob2}

```

### Objective 3
```{r ob3}

```

### Objective 4

#### Objective 4.1
```{r ob4.1}

```

#### Objective 4.2
```{r ob4.2}

```

### GitHub Log
```{bash gitlog} 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```





