---
title: "Group Project: Covid Origins"
author: "Name: Sal Figueroa  \n Partner: Kylie Stearns"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    toc: true
    toc_depth: 3
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
#  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}
---
```{r setup, include = FALSE}

#### Load necessary packages ####
# * These packages are not necessary to complete the assignment and or only used 
#   to provide an example. 
packages <- c("knitr", "kableExtra", "magrittr", "readr", "geosphere")

install_me <- packages[!(packages %in% installed.packages()[, "Package"])]
if (length(install_me)) install.packages(install_me)

library(knitr)
#library(kableExtra)
library(magrittr)
library(readr)
library(geosphere)
```

```{r Data setup, include = FALSE}
#Temp load line from local drive
#time_series_covid19_confirmed_global.csv
confirmed_global <- read.csv("X:\\OneDrive\\CSIT165_RProgramming\\_MasterRepos\\COVID-19\\csse_covid_19_data\\csse_covid_19_time_series\\time_series_covid19_confirmed_global.csv")

#time_series_covid19_deaths_global.csv
deaths_global <- as.data.frame(read.csv("https://raw.githubusercontent.com/Figgs0bit/CSIT165-CovidGroupProj/refs/heads/main/time_series_covid19_deaths_global.csv"))

#time_series_covid19_recovered_global.csv
recovered_global <- as.data.frame(read.csv("https://raw.githubusercontent.com/Figgs0bit/CSIT165-CovidGroupProj/refs/heads/main/time_series_covid19_recovered_global.csv"))
```

## Background
The World Health Organization has recently employed a new data science initiative, *CSIT-165*, that uses data science to characterize pandemic diseases. 
*CSIT-165* disseminates data driven analyses to global decision makers.

*CSIT-165* is a conglomerate comprised of two fabricated entities: *Global Health Union (GHU)* and *Private Diagnostic Laboratories (PDL)*. 
Your and your partner's role is to play a data scientist from one of these two entities.

## Data
> [2019 Novel Coronavirus COVID-19 (2019-nCoV) Data Repository by John Hopkins CSSE](https://github.com/CSSEGISandData/COVID-19/tree/master/csse_covid_19_data/csse_covid_19_time_series)

Data for 2019 Novel Coronavirus is operated by the John Hopkins University Center for Systems Science and Engineering (JHU CSSE).
Data includes daily time series CSV summary tables, including confirmations, recoveries, and deaths. 
Country/region are countries/regions hat conform to World Health Organization (WHO).
Lat and Long refer to coordinates references for the user. 
Date fields are stored in MM/DD/YYYY format.

## Project Objectives

### Objective 1
*Predict where the origin started based on the area with the greatest number of confirmations and deaths on the first recorded day in the data set. Show this is the origin using an if statement.*

```{r ob1}
#find max value in column 5 (10.22.20)
#check to see how many max location there is
#run conditional loop 
max_value <- 0
RowMax <- 0

for(i in 1:nrow(deaths_global[,]))
{
num <- deaths_global[i,5]
  
  if (num > max_value)
    {
    max_value <- num
    RowMax <- i
    }
}

#deaths on x1.22.20 (date)
DeathVal <- deaths_global[RowMax,5]
cat("Province/State:",deaths_global[RowMax,1],"Province/State:",deaths_global[RowMax,2],"\n")
cat("Latitude:",deaths_global[RowMax,3],",Longitude:",deaths_global[RowMax,4],"\n")
cat("Predicted area of Origin had the most initial deaths (01.22.2020) at:",deaths_global[RowMax,5],"\n")
```

### Objective 2
*Where is the most recent area to have a first confirmed case? To do this, you will need to use a for loop, if statement, and subsets.*

```{r ob2}
#find max value in column 5 (10.22.20)
#check to see how many max location there is
#run conditional loop 
max_value <- 0
RowMax <- 0
num <- 0
i <- 5
j <- 1

while(j < 331484 && num == 0) #while(num == 0 && j < 290)
 {
  num <- confirmed_global[j,i]
#  cat("INSIDE WHILE: j:",j,"i:",i,"\n")
  
  j <- j + 1  #Iterate through rows 1-289 

     if( j %% 289 == 0)
      {
      i <- i + 1 #Iterate through columns 1-1147
#      cat("INSIDE IF - Col: ",j,"Row",i,"\n")
      }  
}

j <- j - 1 # 

cat("Province/State: ",confirmed_global[j,1],"Province/State: ",confirmed_global[j,2],"\n")
cat("Latitude: ",confirmed_global[j,3]," ,Longitude:",confirmed_global[j,4],"\n")
cat("Predicted area of Origin had the most initial deaths (01.22.2020) at:",confirmed_global[j,i],"\n")

```


### Objective 3
*How far away are the areas from objective 2 from where the first confirmed case(s) occurred? Please provide answer(s) in terms of miles. Use the function distm from the R package geosphere to calculate the distance between two coordinates in meters (geosphere::distm). You will need to convert the value returned by distm from meters to miles (this conversion is simple and can be found online). Please use a table or printed statement to describe what Province/State and Country/Region first confirmed cases occurred as well as the distance (in miles) away from the origin. Please print the following: {recent region} is {distance in miles} away from {origin city, origin country}.*

```{r ob3}

```

### Objective 4
*CSIT-165 characterizes diseases using risk scores. Risk scores are calculated as the ratio of deaths to confirmations, that is Riskscore=100×deathsconfirmations. Risk scores equal to 100 indicate the highest risk while risk scores equal to 0 indicate the lowest risk. Areas are characterized as being especially vulnerable to loss if they have higher risk scores. For this assignment, exclude cruise ships (hint: they have lat and long coordinates of 0 or NA in this data set, filter this out before calculating risk scores).*

*Which area of the world currently has the lowest risk score (if more than one, display the one with the most confirmations)? Which area of the world currently has the highest risk score (if more than one, display the one with the most confirmations)? How do risk scores in these areas compare to global risk score? Why might it be helpful to calculate metrics like risk scores for different areas of the world and what would their limitations be (what assumptions does risk score make and what important variables might be left out)?*

#### Objective 4.1
```{r ob4.1}

```

#### Objective 4.2
```{r ob4.2}

```


#### Objective 5
*You are asked to make two tables with the top 5 countries that have the most COVID-19 related confirmations and and deaths. Make sure to include all of the counts for the country, not just the counts for one area in the country. To do this we will need to sum all of the values for each country, create new data frames from these values, and use the package kable to convert those data frames into tables.*

*Hint: Sum each country’s counts by subsetting the data frame using a list of countries available in the data set. Use a for loop to iterate through the data frame using the list of countries. For each country, calculate the count sum and assign this value to a list.*
```{r ob5}

```


### GitHub Log
```{bash gitlog} 
git log --pretty=format:"%nSubject: %s%nAuthor: %aN%nDate: %aD%nBody: %b"
```





